---
title: "NEWS"
knitr:
  opts_chunk:
    collapse: true
    comment: "#>"
---

```{r}
#| label: pkgs
#| include: false

repo <- "https://github.com/tidywf/tidydragen"
child1 <- system.file(
  "doc-templates/vignettes/_news.qmd",
  package = "nemo"
)
```

```{r, child=child1}
#| label: child-news
```

## v0.0.0.9000 (dev)

Initial development release. `tidydragen` is a [nemo][nemo] child package that
parses and tidies outputs from Illumina DRAGEN pipelines into wide-format tidy
tibbles ready for a data warehouse, writing to parquet, TSV, CSV, RDS, or a
database. The following pipelines are supported for starters: DNA tumor-normal
(somatic), DNA germline-only, RNA tumor-only, and ctTSO500 (the DRAGEN TSO500
ctDNA app layer).

### Tools

Six domain tools, each with its own `inst/config/tools/<tool>/schema.yaml`, all
inheriting an intermediate `DragenTool` base (itself `inherit = nemo::Tool`) and
orchestrated by the `Dragen` workflow class (exported `DRAGEN_TOOLS`), which
tolerates absent files so each pipeline can expose a different subset.

- `DragenMap`: mapping/time metrics, fragment-length histogram, and the cttso
  trimmer / UMI / GC-bias metrics.
- `DragenFqc`: the 8 flat per-section FastQC tables.
- `DragenCov`: coverage metrics/bins/cumulative, contig mean, fine histogram,
  and cov-report-bed tables across every region
  (`wgs`/`tmb`/`exon`/`target_bed`/ `qc-coverage-region-*`) and phenotype.
- `DragenVar`: variant/SV/CNV/ploidy/TMB/HRD/gVCF metrics plus microsatellite
  (JSON), native ploidy-VCF parse, and contamination (JSON).
- `DragenRna`: fusion and quant metrics.
- `DragenTso`: the ctTSO app-layer (`r pr(3)`).

### The DRAGEN metrics parser

Most `*_metrics.csv` files share a headerless `section,rg,variable,count[,pct]`
shape. `DragenTool` registers a `dragen-metrics` ftype: a plain metrics table is
parsed and pivoted wide (one column per mapped metric plus a paired
`<metric>_pct`) with no per-table code; only tables needing a custom split or a
region/chrom `normalise` hook declare explicit `tidy_<table>()` one-liners.
Fail-loud policies (`on_unmapped`, `on_coerce_fail`, `on_unexpected_col`) catch
schema drift instead of silently dropping data.

### ctTSO app-layer support

- `r pr(3)`

`DragenTso` parses the TSO500 ctDNA outputs beyond the shared DRAGEN metrics:
CombinedVariantOutput small variants, Fusions, TMB trace/msaf, exon/gene
coverage, and a 6-way fan-out of the `SampleAnalysisResults.json`.

### Test data

- `r pr(8)`

Test fixtures rebuilt from real DRAGEN pipeline outputs (trimmed per file), with
real sample IDs scrubbed to `sampleA`/`sampleB`, and tracked with
[DVC](https://dvc.org) on a Cloudflare R2 remote. The default remote is
public-read, so a fresh clone or CI run fetches fixtures with `dvc pull` and no
credentials. The `@testexamples` assertions were updated to the real-data
values; the roxytest suite passes.

### Infrastructure

- CLI, conda recipe, and multi-arch (amd64 + arm64) Docker image (`r pr(6)`).
- GitHub Actions for version bumping and deploy, refactored to the reusable
  `tidywf/actions` workflows and tag-triggered (`r pr(4)`, `r pr(5)`).
- Vignettes: installation, quickstart, structure, schema table, output naming,
  PostgreSQL, CI/CD, and UML.

[nemo]: https://github.com/tidywf/nemo
