---
title: "Output Naming"
knitr:
  opts_chunk:
    collapse: true
    comment: "#>"
---

```{r}
#| label: pkgs
#| include: false

{
  use("nemo")
  use("tidydragen")
  use("fs")
  use("arrow", "read_parquet")
  use("purrr", "map")
  use("dplyr")
  use("glue", "glue")
}
set.seed(1)

# generic template comes from nemo; the special cases below are tidydragen-only
child1 <- system.file(
  "doc-templates/vignettes/_output_naming.qmd",
  package = "nemo"
)
ex <- system.file("extdata", package = "tidydragen")
```

```{r, child=child1}
#| label: child-output-naming
```

## Special cases: region and phenotype prefixes

DRAGEN coverage emits the same table per region (`wgs`, `tmb`, `exon`,
`target_bed`, `--qc-coverage-region-N`) and, for tumor-normal, per phenotype.
These strip to the same prefix (`sampleA` / `sampleA_2`). Subclasses override
the private `refine_files()` hook (runs *before* disambiguation) to put the
region/phenotype in the prefix. Remaining `_2`/`_3` = repeat runs.

### DragenCov: region + phenotype folded into the prefix

- Problem: `sampleA.{wgs,target_bed,exon}_coverage_metrics.csv` all strip to
  `sampleA` via the same parser; somatic `..._normal.csv`/`..._tumor.csv`
  contig-mean files collide too.
- Fix: prefix becomes `sampleA_<region>[_<pheno>]`, e.g.
  `sampleA_wgs_dragencov_metricsmain`, `sampleA_wgs_normal_...`.

```{r}
#| label: cov-init

dir_inC <- path(ex, "dragencov")
dir_tree(dir_inC)
cv <- DragenCov$new(path = dir_inC)
```

One file per region (and phenotype):

```{r}
#| label: cov-run

dir_outC <- dir_create(path(tempdir(), "cov-out"))
cv$run(
  input_id = "input1",
  output_id = "output1",
  output_dir = dir_outC,
  format = "parquet",
  prefix_include = TRUE
)
list.files(dir_outC, pattern = "\\.parquet$") |> sort()
```

`input_prefix` carries the same distinction:

```{r}
#| label: cov-merge

dir_ls(dir_outC, regexp = "dragencov_contigmean\\.parquet") |>
  purrr::map(\(x) arrow::read_parquet(x) |> dplyr::slice_head(n = 2)) |>
  purrr::list_rbind() |>
  dplyr::select(input_id, input_prefix, output_id, dplyr::everything()) |>
  dplyr::slice_head(n = 6)
```

### Variant calling: region kept as a column

`DragenVar` `vc` metrics keep region (`genome`/`targetreg`/`qccovreg`) as a
`region` column (private `region_split()`), since one file holds several
regions: a within-file dimension, not a between-file collision.

### When to reach for the hook

- Prefer schema `pattern`s that already separate variants
- Use `refine_files()` when one parser matches files needing a real label, not
  `_2` (DragenCov)
- Use a column when it varies *within* a file (DragenVar `vc`)
