---
title: "Output Naming"
knitr:
  opts_chunk:
    collapse: true
    comment: "#>"
---

```{r}
#| label: pkgs
#| include: false

{
  use("nemo")
  use("tidydragen")
  use("fs")
  use("arrow", "read_parquet")
  use("purrr", "map")
  use("dplyr")
  use("glue", "glue")
}
set.seed(1)

# generic template comes from nemo; the special cases below are tidydragen-only
child1 <- system.file(
  "doc-templates/vignettes/_output_naming.qmd",
  package = "nemo"
)
ex <- system.file("extdata", package = "tidydragen")
```

```{r, child=child1}
#| label: child-output-naming
```

## Special cases: region and phenotype prefixes

The disambiguation above (`_2`, `_3`) keeps files from overwriting each other
but tells you nothing about *why* two inputs collided. DRAGEN's coverage stage
emits the same logical table for several genomic regions (`wgs`, `tmb`, `exon`,
`target_bed`, and one per user-defined `--qc-coverage-region-N`), and, in the
somatic tumor-normal pipeline, once per phenotype (`normal` / `tumor`). After
the schema `pattern` is stripped these would collapse to the same prefix and
overwrite one another (or land as a lossy `sampleA_2`), so you could no longer
tell which region or phenotype a file came from.

A `Tool` subclass fixes this by overriding the private `refine_files` hook,
which runs on the base prefix *before* the generic disambiguation passes. So a
subclass can fold a meaningful distinction into the prefix itself. Once each
region/phenotype carries a different prefix they no longer collide, and any
remaining `_2`/`_3` is a genuine repeat-run marker.

### DragenCov: region + phenotype folded into the prefix

- Problem: `sampleA.wgs_coverage_metrics.csv`,
  `sampleA.target_bed_coverage_metrics.csv`, and
  `sampleA.exon_coverage_metrics.csv` all strip to the prefix `sampleA` and all
  route through the same coverage parser. In a somatic run the per-phenotype
  contig-mean files (`..._normal.csv` / `..._tumor.csv`) collide too.
- Fix: the `DragenCov` hook rewrites the prefix to `sampleA_<region>[_<pheno>]`
  *before* the collision check, so `sampleA.wgs_coverage_metrics.csv` becomes
  `sampleA_wgs_dragencov_metricsmain` and the normal/tumor contig-mean files
  become `sampleA_wgs_normal_...` / `sampleA_wgs_tumor_...`, never a lossy
  `sampleA_2`.

Point a `DragenCov` object at the bundled coverage fixtures:

```{r}
#| label: cov-init

dir_inC <- path(ex, "dragencov")
dir_tree(dir_inC)
cv <- DragenCov$new(path = dir_inC)
```

Now tidy it. Each region (and, for the contig-mean tables, each phenotype) lands
in its own file, distinguished by the prefix rather than a positional `_2`:

```{r}
#| label: cov-run

dir_outC <- dir_create(path(tempdir(), "cov-out"))
cv$run(
  input_id = "input1",
  output_id = "output1",
  output_dir = dir_outC,
  format = "parquet",
  prefix_include = TRUE
)
list.files(dir_outC, pattern = "\\.parquet$") |> sort()
```

Note how the region (and phenotype where present) is carried in the filename
prefix, and the `input_prefix` column records the same distinction inside each
table:

```{r}
#| label: cov-merge

dir_ls(dir_outC, regexp = "dragencov_contigmean\\.parquet") |>
  purrr::map(\(x) arrow::read_parquet(x) |> dplyr::slice_head(n = 2)) |>
  purrr::list_rbind() |>
  dplyr::select(input_id, input_prefix, output_id, dplyr::everything()) |>
  dplyr::slice_head(n = 6)
```

### Variant calling: region kept as a column

`DragenVar`'s variant-caller metrics (`vc`) take the opposite approach: rather
than folding the region into the prefix, they keep it as an explicit `region`
column via the private `region_split()` helper, since a single file legitimately
carries metrics for more than one region (`genome` / `targetreg` / `qccovreg`).
This is the right call when the distinction is a *within-file* dimension rather
than a *between-file* collision.

### When to reach for the hook

- Prefer schema `pattern`s that already separate variants where you can; the
  generic pipeline then needs no help.
- Use `refine_files()` when a single parser matches multiple input files that
  must stay apart with a meaningful label rather than a positional `_2`
  (DragenCov region/phenotype).
- Keep the distinction as a column when it varies *within* a single file
  (DragenVar `vc` region).
