---
title: "Quickstart"
knitr:
  opts_chunk:
    collapse: true
    comment: "#>"
---

```{r}
#| label: setup
#| include: false

use("tidydragen")
use("arrow", "read_parquet")
use("fs", "dir_tree")
options(width = 90)
```

## Test data

`inst/extdata/<tool>/`; larger files tracked with [DVC](https://dvc.org/) on a
public R2 bucket. `dvc pull` in a clone, or from R:

```r
input_dir <- system.file("extdata/dragencov", package = "tidydragen")
nemo::dvc_download_all(input_dir, file.path(tempdir(), "dvc_test"))
```

## Input

```{r}
#| label: input_tree
#| code-fold: true
#| code-summary: "View input files"

indir <- system.file("extdata", package = "tidydragen")
dir_tree(indir, invert = TRUE, glob = "*.dvc")
```

## Output - Single Tool

Each tool has its own R6 class. `run()` parses, tidies and writes in one step:

```{r}
#| label: single_tool
#| message: false

outdir <- file.path(tempdir(), "qs_map")
DragenMap$new(file.path(indir, "dragenmap"))$run(
  output_dir = outdir,
  format = "parquet",
  input_id = "sampleA_id"
)
list.files(outdir, pattern = "\\.parquet$")
```

### File naming

`{prefix}_{tool}_{table}.parquet`, prefix from input filename (`sampleA`). One
file can fan out to several tables. See [Output Naming](output_naming.html).

### Reading a table back

```{r}
#| label: read_back

m <- read_parquet(file.path(outdir, "sampleA_dragenmap_metrics.parquet"))
m |> str(list.len = 10)
```

## Output - Full DRAGEN run

`Dragen` runs all tools on a parent directory; absent tools are skipped, so it
works for germline, somatic, RNA and ctTSO500:

```{r}
#| label: dragen_run
#| message: false

outdir_d <- file.path(tempdir(), "qs_dragen")
d <- Dragen$new(indir)
d$run(
  output_dir = outdir_d,
  format = "parquet"
)
list.files(outdir_d, pattern = "\\.parquet$") |> sort() |> str()
```

## ID columns

Optional, off by default:

| Argument                | Column added   | Contains                                   |
| ----------------------- | -------------- | ------------------------------------------ |
| `input_id = "x"`        | `input_id`     | sample or run identifier you supply        |
| `output_id = "x"`       | `output_id`    | processing run identifier you supply       |
| `prefix_include = TRUE` | `input_prefix` | filename prefix extracted from input files |

```{r}
#| label: id_cols
#| message: false

outdir_id <- file.path(tempdir(), "qs_map_id")
DragenMap$new(file.path(indir, "dragenmap"))$run(
  output_dir = outdir_id,
  format = "parquet",
  input_id = "sampleA",
  output_id = "out1",
  prefix_include = TRUE
)
read_parquet(file.path(outdir_id, "sampleA_dragenmap_metrics.parquet")) |>
  str(list.len = 10)
```

## Metadata

Every `run()` also writes metadata (input/output dirs, IDs, R package versions):
`metadata.parquet` for `Dragen`, `metadata_<tool>.parquet` for a single tool.

```{r}
#| label: metadata

read_parquet(file.path(outdir_d, "metadata.parquet")) |> str()
```

## See also

- [Schema table](schema_table.html): all tables and columns
- [Structure](structure.html): schemas, versioning, class hierarchy
- [PostgreSQL](postgresql.html)
