---
title: "Quickstart"
knitr:
  opts_chunk:
    collapse: true
    comment: "#>"
---

```{r}
#| label: setup
#| include: false

use("tidydragen")
use("arrow", "read_parquet")
use("fs", "dir_tree")
options(width = 90)
```

tidydragen turns raw [Illumina
DRAGEN](https://www.illumina.com/products/by-type/informatics-products/dragen-secondary-analysis.html)
output directories into consistently structured, versioned, analysis-ready
tables/files. This vignette walks through the core usage patterns.

## Test data

Example input files ship in `inst/extdata/`, one subdirectory per tool. Larger
real-world DRAGEN results are tracked separately via [DVC](https://dvc.org/) and
can be downloaded with either:

- `dvc pull`: requires dvc installation and cloned tidydragen source repo.
- `?nemo::dvc_download_all()`: no additional requirements, uses
  `download.file()` internally.

No credentials are required since the remote is a public Cloudflare R2 bucket.

## Input

Example DRAGEN results (one subdirectory per tool):

```{r}
#| label: input_tree
#| code-fold: true
#| code-summary: "View input files"

indir <- system.file("extdata", package = "tidydragen")
dir_tree(indir)
```

## Output - Single Tool

Each tool has its own R6 class. Construct it with the tool's output directory,
then call `run()` to parse, tidy, and write in one step:

```{r}
#| label: single_tool
#| message: false

outdir <- file.path(tempdir(), "qs_map")
DragenMap$new(file.path(indir, "dragenmap"))$run(
  output_dir = outdir,
  format = "parquet",
  input_id = "sampleA_id"
)
list.files(outdir, pattern = "\\.parquet$")
```

### File naming

Output files follow the pattern `{prefix}_{tool}_{table}.parquet`, where
`prefix` is derived from the input filenames (here `sampleA`). A single input
file can fan out into several tables (e.g. the mapping outputs yield `metrics`,
`time`, `fraglenhist`, and more). The `metadata.parquet` file is always written
alongside the data files and records input/output paths and package versions.

### Reading a table back

```{r}
#| label: read_back

m <- read_parquet(file.path(outdir, "sampleA_dragenmap_metrics.parquet"))
m |> str()
```

## Output - Full DRAGEN run

`Dragen` processes all supported tools in one call. Point it at a directory
containing DRAGEN results; tools whose files are absent simply contribute
nothing, so the same call works across e.g. the germline, somatic tumor-normal,
RNA and ctTSO500 pipelines:

```{r}
#| label: dragen_run
#| message: false

outdir_d <- file.path(tempdir(), "qs_dragen")
d <- Dragen$new(indir)
d$run(
  output_dir = outdir_d,
  format = "parquet"
)
list.files(outdir_d, pattern = "\\.parquet$") |> sort() |> str()
```

## ID columns

Three optional columns can be prepended to every written table. All are off by
default:

| Argument                | Column added   | Contains                                   |
| ----------------------- | -------------- | ------------------------------------------ |
| `input_id = "x"`        | `input_id`     | sample or run identifier you supply        |
| `output_id = "x"`       | `output_id`    | processing run identifier you supply       |
| `prefix_include = TRUE` | `input_prefix` | filename prefix extracted from input files |

These are useful when loading results from multiple samples into the same
database table or combined data frame:

```{r}
#| label: id_cols
#| message: false

outdir_id <- file.path(tempdir(), "qs_map_id")
DragenMap$new(file.path(indir, "dragenmap"))$run(
  output_dir = outdir_id,
  format = "parquet",
  input_id = "sampleA",
  output_id = "out1",
  prefix_include = TRUE
)
read_parquet(file.path(outdir_id, "sampleA_dragenmap_metrics.parquet")) |>
  str()
```

## Metadata

Every `run()` writes a `metadata.parquet` alongside the data files. It records
the input directory, output directory, IDs, and the versions of R packages used:

```{r}
#| label: metadata

read_parquet(file.path(outdir_d, "metadata.parquet")) |> str()
```

## Other useful articles

- [Schema table](schema_table.html): browse every table and column for all
  supported DRAGEN tools
- [Structure](structure.html): how schemas, versioning, and the Tool/Workflow
  class hierarchy work
- [Output naming](output_naming.html): how region/phenotype variants are folded
  into the output prefix
- [PostgreSQL](postgresql.html): writing results to a database
